\documentclass[11pt]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{tikz}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}

\title{\textbf{Formula 1 Ratings and Market Efficiency}}
\author{Peter Cotton}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent We compare rating systems on the full history of Formula 1
(1{,}158 races, 1950 to present). Grids of up to fifty-five cars break the
pairwise decompositions used by standard systems: Glicko-2 scores below a
uniform baseline. A Thurstonian lattice system, which computes joint order
statistics exactly and accepts any performance noise distribution,
outperforms TrueSkill, Elo, Glicko-2 and OpenSkill, and improves further
when its noise density includes a separated block of slow mass matching the
retirement process. We then attempt the harder task of testing market
efficiency with the incomplete odds record that survives in web archives
(135 races). The market leads the best ratings by 0.57 of log loss and a
fitted combination gives the ratings zero weight.
\end{abstract}

\section{Introduction}

A rating system assigns each competitor a skill estimate, updates it after
every contest, and predicts future outcomes. Most systems in use were
designed for two-player games and extend to multi-entrant events by
decomposition: Elo, Glicko-2 and the Weng--Lin systems treat a finish order
as a collection of pairwise results, and TrueSkill chains adjacent
finishers. For small fields the decomposition costs little. Formula 1
provides fields of ten to fifty-five cars, where a single race implies up
to 190 correlated pairwise comparisons, and we will see that the standard
decompositions do not merely lose accuracy there but can fail outright.
The primary purpose of this paper is to compare rating systems on this
terrain and to explain why a Thurstonian treatment, meaning one that works
with the joint distribution of the full finish order, is essential to
doing well on it.

Formula 1 also poses an unusual distributional problem. The fraction of
entrants failing to finish, by decade:

\begin{center}
\begin{tabular}{lcccccccc}
\toprule
decade & 1950s & 60s & 70s & 80s & 90s & 2000s & 10s & 20s \\
\midrule
DNF rate & .42 & .49 & .51 & .59 & .51 & .30 & .18 & .13 \\
\bottomrule
\end{tabular}
\end{center}

\noindent For most of the sport's history, retirement was the most likely
outcome of starting a race. A rating model committed to Gaussian
performance noise has no way to represent this; a model whose noise
density is a free input does. Section 4 exploits this freedom.

A secondary question is whether the resulting ratings can beat the betting
market. We emphasize that with the data available this is a hard test to
run at all: no public archive of Formula 1 winner odds exists, and the
record we recover from web archives is incomplete and unevenly timed.
Section 6 assembles that record and Section 7 reports the comparison,
which favors the market decisively.

\section{Data and protocol}

Race results come from the community-maintained f1db compilation (CC-BY),
covering 1{,}158 championship races from 1950 through the 2026 season.
Drivers completing at least 90\% of the race distance receive official
positions even if they stop early. Other retirements carry no position and
enter our protocol tied at the back of the field. Shared drives in the
1950s produce genuine dead heats, which are treated as ties.

Evaluation is prequential. Races are processed in date order; the first
20\% are unscored warm-up; each later race is predicted before it is
observed. We report log loss, Brier score and accuracy for the winner
prediction, Kendall's tau between predicted and observed orders, expected
calibration error (ECE) over pooled win probabilities, and a rank-PIT
statistic that measures whether the predicted distribution over finish
positions matches the realized positions.

\section{Rating systems compared}

The Thurstonian system used here is built on a discrete lattice of ability
values. Each driver's belief is a probability density on the lattice,
unrestricted by any parametric family. A race performance is ability plus
noise, with the noise drawn from a base density that is an input to the
system. Between races, beliefs diffuse in proportion to elapsed time.

Prediction is a forward computation: each driver's predictive performance
density is the belief convolved with the base density, and win
probabilities are order statistics of the field, computed exactly on the
lattice, ties included. Updating is a backward computation: given an
observed finish order, the likelihood of the whole order as a function of
one driver's ability is computed exactly, holding opponents at their
predictive densities. Index finishers from fastest to slowest. A forward
pass computes, at each performance level, the probability that finishers
$1$ through $k-1$ performed better in the observed order; a backward pass
does the same for finishers $k+1$ through $N$ performing worse. The
likelihood for the $k$-th finisher is the base density correlated with the
product of the two passes. The update costs $O(N)$ lattice operations per
race and was validated against brute-force simulation.

The benchmark below runs all systems over identical data under the
protocol of Section 2. TrueSkill is configured with the empirical
dead-heat rate, and each package uses its own prediction formula.

\begin{center}
\begin{tabular}{lccccc}
\toprule
system & log loss & accuracy & tau & ECE & rank-PIT \\
\midrule
Thurstone lattice & \textbf{2.225} & \textbf{.316} & \textbf{.328} & .015 & .176 \\
TrueSkill & 2.346 & .275 & .300 & .011 & .199 \\
Elo (multi-entrant) & 2.465 & .303 & .319 & .026 & .220 \\
OpenSkill Plackett--Luce & 2.705 & .259 & .283 & .023 & .223 \\
OpenSkill Thurstone--Mosteller & 2.951 & .007 & .106 & .010 & .628 \\
uniform baseline & 3.162 & .043 & --- & --- & --- \\
Glicko-2 & 3.727 & .271 & .255 & .022 & .166 \\
\bottomrule
\end{tabular}
\end{center}

The joint treatment leads on log loss, accuracy and rank correlation. The
failures are the instructive part. Glicko-2 scores below the uniform
baseline: converting a twenty-car grid into nineteen results within one
rating period destabilizes its volatility estimation, and our
implementation already includes stabilizing guards without which it does
not complete the dataset. OpenSkill's all-pairs Thurstone--Mosteller
variant diverges the same way. These systems are sound at head-to-head
scale; what breaks them here is treating 190 correlated comparisons as
independent evidence. This is why exact joint order statistics are not a
refinement in this setting but a requirement.

\section{Choosing the noise distribution}

Every comparison system above is committed to Gaussian beliefs with
Gaussian or Gumbel noise. The lattice system accepts any base density, so
the noise model becomes an empirical question, and the retirement table in
Section 1 suggests the answer should not be Gaussian.

The motivated candidate is a two-component density: with probability
$1-p$ the noise is standard Gaussian, and with probability $p$ it is drawn
uniformly from a wide interval on the slow side. Figure~\ref{fig:slab}
shows the shape. The rectangular block represents retirement: a mechanical
failure produces a result unrelated to pace, so its probability mass is
separated from the pace distribution rather than blended into its tail.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.9]
  % axes
  \draw[->] (-3.2,0) -- (9.6,0) node[below right] {performance (slower $\rightarrow$)};
  \draw[->] (-3.2,0) -- (-3.2,2.6) node[left] {density};
  % gaussian core (1-p)
  \draw[thick,domain=-3:3,samples=80] plot (\x,{2.2*exp(-0.5*\x*\x)});
  \node at (0,2.45) {Gaussian pace, mass $1-p$};
  % slab
  \draw[thick] (4,0) -- (4,0.55) -- (8.8,0.55) -- (8.8,0);
  \node at (6.4,0.85) {retirement block, mass $p$};
  % ticks
  \draw (0,0.06) -- (0,-0.06) node[below] {$0$};
  \draw (4,0.06) -- (4,-0.06) node[below] {$3$};
  \draw (8.8,0.06) -- (8.8,-0.06) node[below] {$12$};
\end{tikzpicture}
\caption{The disaster-slab base density: a continuous Gaussian pace
component and a separated rectangular block of slow mass representing
retirement.}
\label{fig:slab}
\end{figure}

\begin{center}
\begin{tabular}{lccc}
\toprule
base density & log loss & ECE & rank-PIT \\
\midrule
era-adaptive slab & \textbf{2.200} & .016 & .185 \\
$N(0,1)$ plus 25\% slab & 2.203 & .015 & .158 \\
skew-normal ($a{=}2$) & 2.222 & \textbf{.010} & .165 \\
Gaussian & 2.225 & .015 & .176 \\
tight core (0.5) plus 25\% slab & 2.254 & .013 & \textbf{.136} \\
Student-$t$ ($\nu{=}5$) & 2.271 & .018 & .171 \\
near-Dirac core plus slab & 2.42--2.52 & --- & --- \\
\bottomrule
\end{tabular}
\end{center}

The slab improves every metric relative to the Gaussian. Because a fixed
25\% mass is wrong for most decades, the era-adaptive variant estimates
$p$ from a trailing window of observed retirement frequency, using past
races only. It attains the best log loss and rank correlation of any
configuration tested, and its fitted mass declines across the dataset to
finish near the modern retirement rate.

The failed alternatives are informative. Symmetric heavy tails (the $t$
family) hurt increasingly as tails fatten, because symmetry places equal
mass on implausibly fast performances. Skew-normal noise changes little,
consistent with two facts about rank data: the difference of two
independent identically distributed noises is symmetric whatever their
skew, and ranks are invariant to monotone transformations of the time
scale. A near-Dirac pace core loses about 0.2 of log loss, showing that
variation from strategy, traffic and weather is comparable in size to
ability differences. None of these choices is available to a system
committed to a parametric family, which is the second reason the
Thurstonian treatment matters here.

\section{Summary of the ratings comparison}

Two properties carried the result: exact joint order statistics, without
which systems degrade or diverge at grid scale, and a free noise density,
which converts knowledge of the retirement process into forecast accuracy.
Both are structural. They cannot be patched onto a pairwise Gaussian
system by tuning.

\section{The odds record}

We know of no public archive of Formula 1 winner odds, so we assembled
what survives. The Wayback Machine holds captures of odds-comparison
pages; 523 usable captures yield full-field win odds for 144 races between
2008 and 2025, each covering roughly twenty bookmakers. For each race we
take the median implied probability per driver across bookmakers and
normalize within the race, which removes the bookmaker margin. The record
is incomplete and unevenly timed: 2011--2012 and 2017--2020 are sparse,
many captures postdate qualifying, and a few precede the race by more than
a week. These limitations should be kept in view; they make the efficiency
test below possible but imperfect. We publish only the derived consensus
probabilities.

\section{The efficiency test}

On the 135 covered races falling after the warm-up period, with market and
ratings scored on the same matched set of drivers:

\begin{center}
\begin{tabular}{lc}
\toprule
 & log loss \\
\midrule
market consensus & 1.547 \\
fitted pool ($a{=}1.10$, $b{=}0.00$) & 1.538 \\
ratings given all past odds & 1.898 \\
era-adaptive slab ratings & 2.117 \\
uniform & 3.018 \\
\bottomrule
\end{tabular}
\end{center}

The market leads the best ratings by $0.570$ of log loss (standard error
$0.065$). A pool of the form $q \propto m^a s^b$, with exponents fitted on
past races only, assigns the ratings weight $b = 0$. The result survives
enriching the challenger: feeding the ratings the consensus odds of all
previous races as additional observations improves their standalone score
from 2.117 to 1.898, but the fitted weight remains zero. Prices subsume
everything the ratings know.

The explanation is straightforward. Post-qualifying prices incorporate
grid position, current car performance and weather, none of which is
visible to a system built from past results. The one detectable
miscalibration is the fitted exponent $a = 1.10$, indicating slight
underconfidence about favorites, in the direction of the favorite-longshot
bias documented in other racing markets. It is worth 0.008 of log loss,
well inside any bookmaker margin. Within the limits of the archived
record, Formula 1 winner markets appear efficient with respect to
results-based ratings, including ratings augmented with the market's own
price history.

Two qualifications. First, snapshot timing favors the market, since
post-qualifying information is present in many captures; closing-price
data would sharpen the estimate but is unlikely to reverse a gap of nine
standard errors. Second, the test covers winner markets only. Bookmakers
price derivative markets such as head-to-heads and top-six finishes from
winner odds using simplifying rank models, and our results on rank models
suggest that is where pricing error would concentrate; testing it requires
derivative odds the archives do not contain.

\section{Limitations}

Constructor effects are folded into driver ratings; separating driver from
car is the natural next study. Retirements enter tied at the back, which
discards the official ordering of retirees by laps completed. The ratings
are one-dimensional, with no axis for wet weather or circuit type. The
odds record is as described in Section 6.

\section*{Reproducibility}

Each table is one command from the \texttt{winning} package
(\texttt{github.com/microprediction/winning}):
\texttt{python -m winning.benchmarks.run\_benchmark --dataset f1} for
Section 3, and the scripts \texttt{research/f1\_dirac\_disaster.py},
\texttt{research/t\_sweep.py}, \texttt{research/f1\_era\_slab.py},
\texttt{research/f1\_odds\_fetch.py}, \texttt{research/f1\_odds\_parse.py}
and \texttt{research/f1\_market\_test.py} for Sections 4 through 7. Race
data is fetched at run time from f1db (CC-BY-4.0). The lattice machinery
is the \texttt{thurstone} package; the underlying ability transform is
described in Cotton, \emph{SIAM J.\ Financial Mathematics} 12(1), 2021.

\end{document}
